---
title: "Employee Salary Analysis"
author: "Ken Dafoe"
date: "`r Sys.Date()`"
output:
  html_document:
    css: styles.css
---

```{r Load Data, echo=FALSE, include=FALSE}
library(readxl)
library(janitor)
employee <- read_xlsx("Employee Salaries - ONLY.xlsx")
emp_salaries <- employee
emp_salaries <- clean_names(emp_salaries)
```


```{r New salaries, eval=FALSE, include=FALSE}
# Add 2 new salaries
new_salary1 <- c(22000)
new_salary2 <- c(161000)

# Create a data frame with these new salaries
# new_salaries <- data.frame(annual_salary = c(new_salary1, new_salary2))
# Append new_salaries to emp_salaries
# emp_salaries <- rbind(emp_salaries, new_salaries)

```


```{r Descriptives, include=FALSE}
library(psych)
e_sum <- describe(emp_salaries$annual_salary)
e_sumsal <- sum(emp_salaries)
```


```{r Bin Range, message=FALSE, warning=FALSE, include=FALSE}

# Load the libraries
library(knitr)
library(kableExtra)

# Define the bin width and breaks
bin_width <- 20000
min_salary <- e_sum$min
max_salary <- e_sum$max

bins <- seq(min_salary, max_salary + bin_width, by = bin_width)

# Create bin labels
bin_labels <- paste(formatC(bins[-length(bins)], format = "f", digits = 0, big.mark = ","),
                    formatC(bins[-length(bins)] + bin_width, format = "f", digits = 0, big.mark = ","), sep=" - ")

# Cut the salary data into bins
salary_bins <- cut(emp_salaries$annual_salary, breaks = bins, labels = bin_labels, include.lowest = TRUE)

# Count the frequencies
bin_counts <- table(salary_bins)

# Convert to a data frame
bin_counts_df <- as.data.frame(bin_counts)
names(bin_counts_df) <- c("Salary Range", "Frequency")

# Create a table
knitr::kable(bin_counts_df, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```


```{r Percentile 70th, include=FALSE}
#Sum Frequencies
bottom <- round(sum(bin_counts_df[1:3,"Frequency"])/e_sum[[2]],2)
top <- round(sum(bin_counts_df[-c(1:3),"Frequency"])/e_sum[[2]],2)

#Count of Frequencies
low1 <- sum(bin_counts_df[1:3,"Frequency"])
high1 <- sum(bin_counts_df[-c(1:3),"Frequency"])

# Calculate the 70th percentile of annual_salary
p70 <- quantile(emp_salaries$annual_salary, 0.7)
p70 <- round(p70/1000)*1000


# Sum of annual_salary values less than or equal to the 70th percentile
sum_below_p70 <- round(sum(emp_salaries$annual_salary[emp_salaries$annual_salary <= p70]),2)
pcent_below_p70 <- round(sum_below_p70/e_sumsal*100,0)

# Sum of annual_salary values greater than or equal to the 70th percentile
sum_above_p70 <- round(sum(emp_salaries$annual_salary[emp_salaries$annual_salary >= p70]),2)
pcent_above_p70 <- round(sum_above_p70/e_sumsal*100,0)
```


```{r Inequality, include=FALSE}
# Gini coefficient
library(ineq)
gini_coefficient <- Gini(emp_salaries$annual_salary)

# Employee Counts
emp_count <- length(emp_salaries$annual_salary)
# Calculate the index corresponding to the top 10% and bottom 40% of employees
top_10_index <- round(emp_count * 0.1)
bottom_40_index <- round(emp_count * 0.4)

# Calculate the number of employees at each percentile
num_employees_top_10 <- emp_count - top_10_index + 1
num_employees_bottom_40 <- bottom_40_index

# Calculate the Palma Index
palma_index <- function(income_vector) {
  sorted_income <- sort(income_vector)
  n <- length(sorted_income)
  top_10_percent <- quantile(sorted_income, 0.9)
  bottom_40_percent <- quantile(sorted_income, 0.4)
  
  top_10_income <- sum(sorted_income[sorted_income >= top_10_percent])
  bottom_40_income <- sum(sorted_income[sorted_income <= bottom_40_percent])
  
  palma <- top_10_income / bottom_40_income
  return(palma)
}

palma_i <- palma_index(emp_salaries$annual_salary)
```


```{r Outliers, include=FALSE}
z_scores <- scale(emp_salaries$annual_salary)
outliers <- which(abs(z_scores) > 3)
outliers

Q1 <- quantile(emp_salaries$annual_salary, 0.25)
Q3 <- quantile(emp_salaries$annual_salary, 0.75)
IQR <- Q3 - Q1
IQR_outliers <- which(emp_salaries$annual_salary < Q1 - 1.5 * IQR | emp_salaries$annual_salary > Q3 + 1.5 * IQR)
IQR_outliers

# Geometric Mean of annual_salaries
geometric_mean <- round(exp(mean(log(emp_salaries$annual_salary)))/1000,0)
```


```{r Distributions, warning=FALSE, include=FALSE}
library(fitdistrplus)

# Extract the salary column
salaries <- emp_salaries$annual_salary

# Remove NA values
salaries <- salaries[!is.na(salaries)]
# Remove Inf and -Inf values
salaries <- salaries[is.finite(salaries)]
# Now try fitting again
# fit_norm <- fitdist(salaries, "norm")

# Fit different candidate distributions to the data
fit_norm <- fitdist(salaries, "norm")
fit_lognorm <- fitdist(salaries, "lnorm")
fit_logis <- fitdist(salaries, "logis")

# Compare the goodness-of-fit using AIC
gof_norm <- gofstat(fit_norm)
gof_lognorm <- gofstat(fit_lognorm)
gof_logis <- gofstat(fit_logis)

# Visualize the goodness-of-fit using density function comparison
denscomp(list(fit_norm, fit_lognorm, fit_logis), main = "Density comparison")
```


```{r Normality Tests, include=FALSE}
# Tests for data normality
# Shapiro-Wilk test
shapiro_test_result <- shapiro.test(emp_salaries$annual_salary)[2]
shapiro_f <- sprintf("%.3f", shapiro_test_result)

# Create function to adjust p-value
adjust_pvalue <- function(pvalue) {
  if (pvalue < 0.001) {
    return(0.001)
  } else {
    return(pvalue)
  }
}

adj_shapiro_f <- adjust_pvalue(shapiro_f)

# Lilliefors (Kolmogorov-Smirnov) normality test
library(nortest)
lillie_test_result <- lillie.test(emp_salaries$annual_salary)[2]
lillie_f <- sprintf("%.3f", lillie_test_result)

adj_kol_f <- adjust_pvalue(lillie_f)
```
<p></p>
<p class="tight-paragraphs">This company has `r e_sum[[2]]` employees.</p>
<p class="tight-paragraphs">The sum of all salaries is $`r round(e_sumsal/1000000,1)`mil.</p>
<p class="tight-paragraphs">The average salary is $`r round(e_sum[[3]]/1000,0)`k.</p>
<p class="tight-paragraphs">The minimum salary is \$`r round(min(emp_salaries$annual_salary)/1000,0)`k.</p>
<p class="tight-paragraphs">The maximum salary is \$`r round(max(emp_salaries$annual_salary)/1000,0)`k.</p>
<p class="tight-paragraphs">Which gives a range of $`r round(e_sum[[10]]/1000,0)`k.</p>
<p class="tight-paragraphs">With a wide dispersion supported by a standard deviation of $`r round(e_sum[[4]]/1000,0)`k.</p><p></p>

*A visualization of the salaries in the histogram shows us two things: One is that the income distribution is positively skewed. This suggests a non-normal distribution supported by a series of fit tests that confirm a log-normal distribution with the lowest AIC of* `r round(gof_lognorm$aic[1],1)` *as the best fit.*
*And two, the histogram shows a majority of* `r sum(bin_counts_df[1:3,"Frequency"])` *employees below the 70th percentile of* $`r round(p70[[1]]/1000,0)`k, *and* `r sum(bin_counts_df[4:7,"Frequency"])` *employees above.*

Two data normality tests, Shapiro-Wilk (p-value <`r adj_shapiro_f`) and Kolmogorov-Smirnov (p-value <`r adj_kol_f`), both display a p-value <.05 confirming that the salary data significantly deviates from a normal distribution.

With this knowledge, we performed a log-normal transformation of the salaries to a geometric mean of \$`r round(exp(mean(log(emp_salaries$annual_salary)))/1000,0)`k, indicating a few high-income earners are driving up the average in the original dataset, even with a outlier count of `r ifelse(length(outliers) == 0, print(0), outliers)`. In this context the median salary of \$`r round(median(emp_salaries$annual_salary)/1000,0)`k is a better representative central value.

```{r Frequencies, echo=FALSE, include=TRUE}
hist(emp_salaries$annual_salary, breaks = 6, freq = T,
     col = "blue", main = "Histogram of Annual Salaries",
     xlab = "", ylab = "")
abline(v = p70, col = "orange", lwd = 2)
text(x = p70, y = max(hist(emp_salaries$annual_salary, breaks = 6, plot = FALSE)$counts), labels = paste0(p70/1000, "k"), adj = c(0,1), col = "orange")
```

In fact, `r pcent_below_p70`% of the sum of all salaries are below \$`r round(p70[[1]]/1000,0)`k, while `r pcent_above_p70`% are above.


*This knowledge indicates a possible inequality in the income distribution of the company, supported by a Lorenz curve showing a moderate bowing in the salaries from the orange line of equality.*

```{r Lorenz curve, echo=FALSE,include=TRUE}

# Load the ineq package
library(ineq)

# Assuming emp_salaries$annual_salary is your vector of salaries
salaries <- emp_salaries$annual_salary

# Compute the Lorenz curve
Lc <- Lc(salaries)

# Plot the Lorenz curve
plot(Lc, main="Lorenz Curve", xlab="Population Percentage", 
     ylab="Cumulative Salary", col = "blue")

# Add the line of perfect equality
abline(0, 1, lty=1, col = "orange")

```

Closer examination of the data shows a Gini coefficient of `r round(gini_coefficient,2)`, indicating a moderate inequality in the distribution of income within the company.
Also, the Palma index scores `r round(palma_i,2)`, which indicates that the top 10% of earners hold `r round(palma_i,2)*100`% of the salaries of the bottom 40% of employees.

To summarize our key findings:

- <p class="tight-paragraphs"> `r e_sum[[2]]` employees earn $`r round(e_sumsal/1000000,1)`mil in income.</p>
- <p class="tight-paragraphs"> The central measure of this income is the median salary of \$`r round(median(emp_salaries$annual_salary)/1000,0)`k.</p>
- <p class="tight-paragraphs"> A Palma index of the salaries shows the top 10% of earners hold `r round(palma_i,2)*100`% of the salaries of the bottom 40% of employees.</p>
- <p class="tight-paragraphs"> Salary inequality is more pronounced below the 40th percentile, since there are `r bottom_40_index` employees in this income percentile, versus `r top_10_index` employees in the top 10% of income earners.</p><p></p>


<font size="4">*It is important to recognize that a more thorough investigation on employee salaries would need additional information about the job titles, departments, and levels of responsibility within the company, as well as the distribution of salaries across these different categories.*</font>


From these findings, we propose the following actions:

**Income inequality investigation**
<p class="tight-paragraphs">Would involve looking at factors such as job roles, years of experience, education level, and performance ratings.</p>

**Salary Review**
<p class="tight-paragraphs">Given the high standard deviation and coefficient of variation, it might be beneficial to review the company's salary structure. Are there valid reasons for the wide range of salaries? Are certain roles underpaid or overpaid relative to the market?</p>

**Salary Banding**
<p class="tight-paragraphs">If it's not already in place. This could help to ensure that employees in similar roles with similar experience and performance levels are paid within a certain range of each other.</p>

**Performance Based Pay**
<p class="tight-paragraphs">If not already in place, consider implementing a performance-based pay system. This could help to ensure that high-performing employees are rewarded appropriately.</p><p></p>

